---
title: "Colorado Retail Sales"
format:
  dashboard:
    theme: cosmo
    embed-resources: true
    nav-buttons:
      - icon: database
        href: https://cdor.colorado.gov/retail-sales-reports
      - icon: github
        href: https://github.com/gtottenSDO/colorado-retail-sales
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
library(tidyverse)
library(arrow)
library(scales)
if (requireNamespace("sdotools", quietly = TRUE)) {
  library(sdotools)
}

# Load pre-built parquet files (run R/collect_data.R to refresh)
state_df <- read_parquet("data/state.parquet")
county_df <- read_parquet("data/county.parquet")
city_df <- read_parquet("data/city.parquet")
county_ind_df <- read_parquet("data/county_industry.parquet")
city_ind_df <- read_parquet("data/city_industry.parquet")
state_ind_df <- read_parquet("data/state_industry.parquet")

# Source data embeds stray line breaks in some industry names
clean_industry <- function(df) mutate(df, industry = str_squish(industry))
county_ind_df <- clean_industry(county_ind_df)
city_ind_df <- clean_industry(city_ind_df)
state_ind_df <- clean_industry(state_ind_df)

# Compute YoY for state
state_yoy <- state_df |>
  arrange(year, month) |>
  mutate(yoy = (retail_sales / lag(retail_sales, 12) - 1) * 100)

# Latest month for headlines
latest <- state_df |> slice_max(date, n = 1)
latest_date <- latest$date

# Years with all 12 months reported (annual views exclude partial years)
complete_years <- state_df |>
  count(year) |>
  filter(n == 12) |>
  pull(year)

# Counties list (excluding "Total" aggregate)
counties <- county_df |>
  filter(county != "Total") |>
  distinct(county) |>
  pull(county) |>
  sort()

cities <- city_df |>
  distinct(city) |>
  pull(city) |>
  sort()

# Geographies with published industry detail (large counties/cities only)
ind_counties <- sort(unique(county_ind_df$county))
ind_cities <- sort(unique(city_ind_df$city))

# ---------------------------------------------------------------------------
# Trailing-12-month (T12) window, derived from the latest month in the data
# ---------------------------------------------------------------------------

t12_end_cur <- latest_date
t12_start_cur <- latest_date %m-% months(11)
t12_end_pri <- latest_date %m-% months(12)
t12_start_pri <- latest_date %m-% months(23)

t12_label_cur <- paste(
  format(t12_start_cur, "%b %Y"),
  "–",
  format(t12_end_cur, "%b %Y")
)
t12_label_pri <- paste(
  format(t12_start_pri, "%b %Y"),
  "–",
  format(t12_end_pri, "%b %Y")
)

assign_period <- function(df) {
  df |>
    mutate(
      period = case_when(
        date >= t12_start_cur & date <= t12_end_cur ~ "current",
        date >= t12_start_pri & date <= t12_end_pri ~ "prior",
        TRUE ~ NA_character_
      )
    )
}

# State T12 total (for county excess benchmark)
state_t12_total_yoy <- county_df |>
  filter(county == "Total", !is.na(retail_sales)) |>
  assign_period() |>
  filter(!is.na(period)) |>
  group_by(period) |>
  summarise(sales = sum(retail_sales, na.rm = TRUE), .groups = "drop") |>
  pivot_wider(names_from = period, values_from = sales) |>
  mutate(state_yoy = (current / prior - 1) * 100) |>
  pull(state_yoy)

# County T12 YoY — all 64 counties
county_t12 <- county_df |>
  filter(county != "Total", !is.na(retail_sales)) |>
  assign_period() |>
  filter(!is.na(period)) |>
  group_by(county, period) |>
  summarise(
    sales = sum(retail_sales, na.rm = TRUE),
    n_months = n(),
    .groups = "drop"
  ) |>
  pivot_wider(names_from = period, values_from = c(sales, n_months)) |>
  mutate(
    # Suppressed/missing months would distort the comparison, so growth is
    # only computed for counties with a full 12 months in both windows
    yoy = if_else(
      n_months_current == 12 & n_months_prior == 12,
      (sales_current / sales_prior - 1) * 100,
      NA_real_
    ),
    excess = yoy - state_t12_total_yoy,
    # Flag small counties (prior-period T12 < $500M) for context
    small_county = sales_prior < 500e6
  )

# Default picks: largest geographies by recent sales, not alphabetical
default_counties <- county_t12 |>
  slice_max(sales_current, n = 5) |>
  pull(county)

# City reporting can lag the statewide series, so rank at the cities' own
# latest month rather than latest_date
default_cities <- city_df |>
  filter(!is.na(retail_sales)) |>
  filter(date == max(date)) |>
  slice_max(retail_sales, n = 5) |>
  pull(city)

# Industries: separate current categories from those retired in the 2022
# NAICS revision, and default to the largest current ones by T12 sales
ind_coverage <- state_ind_df |>
  filter(industry != "Total", !is.na(retail_sales)) |>
  group_by(industry) |>
  summarise(
    last_date = max(date),
    t12_sales = sum(retail_sales[date >= t12_start_cur], na.rm = TRUE),
    .groups = "drop"
  )

industries_active <- ind_coverage |>
  filter(last_date >= t12_start_cur) |>
  arrange(industry) |>
  pull(industry)

industries_legacy <- ind_coverage |>
  filter(last_date < t12_start_cur) |>
  arrange(industry) |>
  pull(industry)

default_industries <- ind_coverage |>
  filter(industry %in% industries_active) |>
  slice_max(t12_sales, n = 8) |>
  pull(industry)

# Core retail industries (post-2022 NAICS, 0% NA rate in county_industry)
core_heatmap_inds <- c(
  "Accommodation",
  "Building Material and Garden Equipment and Supplies Dealers",
  "Clothing, Clothing Accessories, Shoe, and Jewelry Retailers",
  "Food and Beverage Retailers",
  "Food Services and Drinking Places",
  "Furniture, Home Furnishings, Electronics, and Appliance Retailers",
  "Gasoline Stations and Fuel Dealers",
  "General Merchandise Retailers",
  "Health and Personal Care Retailers",
  "Motor Vehicle and Parts Dealers",
  "Sporting Goods, Hobby, Musical Instrument, Book, and Miscellaneous Retailers"
)

# State T12 YoY by industry (benchmark for heatmap excess)
state_ind_t12 <- state_ind_df |>
  filter(industry %in% core_heatmap_inds, !is.na(retail_sales)) |>
  assign_period() |>
  filter(!is.na(period)) |>
  group_by(industry, period) |>
  summarise(sales = sum(retail_sales, na.rm = TRUE), .groups = "drop") |>
  pivot_wider(names_from = period, values_from = sales) |>
  mutate(
    state_yoy = (current / prior - 1) * 100,
    sales_t12 = current
  )

# County × industry T12 YoY with excess vs state industry benchmark
heatmap_county_ind <- county_ind_df |>
  filter(industry %in% core_heatmap_inds, !is.na(retail_sales)) |>
  assign_period() |>
  filter(!is.na(period)) |>
  group_by(county, industry, period) |>
  summarise(sales = sum(retail_sales, na.rm = TRUE), .groups = "drop") |>
  pivot_wider(names_from = period, values_from = sales) |>
  mutate(yoy = (current / prior - 1) * 100) |>
  left_join(
    state_ind_t12 |> select(industry, state_yoy),
    by = "industry"
  ) |>
  mutate(excess = yoy - state_yoy)

# Heatmap caveat that expires once Dec 2025 leaves the T12 window
weld_note <- if (t12_start_cur <= as.Date("2025-12-01") &&
  t12_end_cur >= as.Date("2025-12-01")) {
  " Weld County / Sporting Goods contains an anomalous Dec 2025 spike — interpret with caution."
} else {
  ""
}

# Pass key data to OJS
ojs_define(
  state_data = state_yoy,
  county_data = county_df |>
    filter(county != "Total") |>
    arrange(county, year, month) |>
    group_by(county) |>
    mutate(yoy = (retail_sales / lag(retail_sales, 12) - 1) * 100) |>
    ungroup(),
  city_data = city_df |>
    arrange(city, year, month) |>
    group_by(city) |>
    mutate(yoy = (retail_sales / lag(retail_sales, 12) - 1) * 100) |>
    ungroup(),
  county_ind_data = county_ind_df,
  city_ind_data = city_ind_df,
  state_ind_data = state_ind_df,
  counties_list = counties,
  cities_list = cities,
  ind_counties_list = ind_counties,
  ind_cities_list = ind_cities,
  industries_active_list = industries_active,
  industries_legacy_list = industries_legacy,
  default_counties = default_counties,
  default_cities = default_cities,
  default_industries = default_industries,
  complete_years = complete_years,
  latest_month = format(latest$date, "%B %Y"),
  latest_sales = latest$retail_sales,
  latest_yoy = round(state_yoy |> slice_max(date, n = 1) |> pull(yoy), 1),
  t12_label_cur = t12_label_cur,
  # Pre-aggregated T12 data for Snapshot and Heatmap tabs
  county_t12_data = county_t12,
  state_ind_t12_data = state_ind_t12,
  heatmap_data = heatmap_county_ind,
  state_t12_yoy = round(state_t12_total_yoy, 2)
)
```

```{ojs}
//| output: false

// Shared helpers used across all tabs ------------------------------------

indShort = new Map([
  ["Accommodation", "Accommodation"],
  ["Administrative and Support and Waste Management and Remediation Services", "Admin & Waste Services"],
  ["Agriculture, Forestry, Fishing and Hunting", "Agriculture & Forestry"],
  ["Arts, Entertainment, and Recreation", "Arts & Entertainment"],
  ["Building Material and Garden Equipment and Supplies Dealers", "Building Materials"],
  ["Clothing and Clothing Accessories Stores", "Clothing Stores (thru 2021)"],
  ["Clothing, Clothing Accessories, Shoe, and Jewelry Retailers", "Clothing & Jewelry"],
  ["Electronics and Appliance Stores", "Electronics Stores (thru 2021)"],
  ["Finance and Insurance", "Finance & Insurance"],
  ["Food and Beverage Retailers", "Food & Beverage Retail"],
  ["Food and Beverage Stores", "Food & Bev Stores (thru 2021)"],
  ["Food Services and Drinking Places", "Restaurants & Bars"],
  ["Furniture and Home Furnishings Stores", "Furniture Stores (thru 2021)"],
  ["Furniture, Home Furnishings, Electronics, and Appliance Retailers", "Furniture & Electronics"],
  ["Gasoline Stations", "Gas Stations (thru 2021)"],
  ["Gasoline Stations and Fuel Dealers", "Gas & Fuel"],
  ["General Merchandise Retailers", "General Merchandise"],
  ["General Merchandise Stores", "General Merch Stores (thru 2021)"],
  ["Health and Personal Care Retailers", "Health & Personal Care"],
  ["Health and Personal Care Stores", "Health Stores (thru 2021)"],
  ["Health Care and Social Assistance", "Health Care"],
  ["Management of Companies and Enterprises; Other Services (except Public Administration)", "Management & Other Services"],
  ["Mining, Quarrying, and Oil and Gas Extraction", "Mining, Oil & Gas"],
  ["Miscellaneous Store Retailers", "Misc Retailers (thru 2021)"],
  ["Motor Vehicle and Parts Dealers", "Motor Vehicles"],
  ["Nonstore Retailers", "Nonstore Retail (thru 2021)"],
  ["Professional, Scientific, and Technical Services", "Professional Services"],
  ["Real Estate and Rental and Leasing", "Real Estate & Rental"],
  ["Sporting Goods, Hobby, Musical Instrument, and Book Stores", "Sporting Goods Stores (thru 2021)"],
  ["Sporting Goods, Hobby, Musical Instrument, Book, and Miscellaneous Retailers", "Sporting Goods & Hobbies"],
  ["Transportation and Warehousing", "Transportation & Warehousing"]
])

shortInd = (name) => indShort.get(name) ?? name

fmtUSD = (d) => {
  const abs = Math.abs(d)
  if (abs >= 1e9) return `$${(d / 1e9).toFixed(1)}B`
  if (abs >= 1e6) return `$${(d / 1e6).toFixed(0)}M`
  if (abs >= 1e3) return `$${(d / 1e3).toFixed(0)}K`
  return `$${d.toFixed(0)}`
}

fmtSigned = (d) => `${d > 0 ? "+" : ""}${d.toFixed(1)}%`

completeYears = new Set(complete_years)

legacySet = new Set(industries_legacy_list)

emptyNote = (msg) =>
  html`<div style="padding: 2rem; color: #666; font-style: italic;">${msg}</div>`
```

# State Overview

## Row {height="15%"}

```{r}
#| content: valuebox
#| title: "Data Through"
list(
  icon = "calendar3",
  color = "secondary",
  value = format(latest$date, "%B %Y")
)
```

```{r}
#| content: valuebox
#| title: "Retail Sales (Latest Month)"
list(
  icon = "shop",
  color = "primary",
  value = dollar(
    latest$retail_sales,
    scale = 1e-9,
    suffix = "B",
    accuracy = 0.1
  )
)
```

```{r}
#| content: valuebox
#| title: "vs. Same Month Last Year"
yoy_val <- state_yoy |> slice_max(date, n = 1) |> pull(yoy)
list(
  icon = if (yoy_val >= 0) "arrow-up" else "arrow-down",
  color = if (yoy_val >= 0) "success" else "danger",
  value = paste0(sprintf("%+.1f", yoy_val), "%")
)
```

```{r}
#| content: valuebox
#| title: "Trailing 12-Month Growth"
list(
  icon = if (state_t12_total_yoy >= 0) "graph-up-arrow" else "graph-down-arrow",
  color = if (state_t12_total_yoy >= 0) "success" else "danger",
  value = paste0(sprintf("%+.1f", state_t12_total_yoy), "%")
)
```

## Row

```{ojs}
//| title: "Monthly Retail Sales (Statewide)"

viewof time_unit = Inputs.radio(["Monthly", "Annual"], { value: "Monthly", label: "View:" })

{
  const isAnnual = time_unit === "Annual"
  const data = transpose(state_data).filter(d => d.retail_sales != null && !isNaN(d.retail_sales))
  let plotData = data

  if (isAnnual) {
    const byYear = d3.rollup(
      data.filter(d => completeYears.has(d.year)),
      v => ({ year: v[0].year, retail_sales: d3.sum(v, d => d.retail_sales) }),
      d => d.year
    )
    plotData = [...byYear.values()].sort((a, b) => a.year - b.year)
  }

  return Plot.plot({
    width: 800,
    height: 320,
    x: isAnnual ? { label: null, tickFormat: d => String(d) } : { label: null },
    y: { label: "Retail sales", tickFormat: fmtUSD },
    caption: isAnnual
      ? "Complete calendar years only"
      : `Source: Colorado Dept. of Revenue · Data through ${latest_month}`,
    marks: [
      Plot.areaY(plotData, {
        x: isAnnual ? "year" : (d => new Date(d.date)),
        y: "retail_sales",
        fill: "#2E6DA4",
        fillOpacity: 0.2
      }),
      Plot.lineY(plotData, {
        x: isAnnual ? "year" : (d => new Date(d.date)),
        y: "retail_sales",
        stroke: "#2E6DA4",
        strokeWidth: 2,
        tip: { format: { y: fmtUSD } }
      })
    ]
  })
}
```

```{ojs}
//| title: "Year-over-Year Growth (vs. Same Month, Prior Year)"

{
  const data = transpose(state_data).filter(d => d.yoy != null && !isNaN(d.yoy))
  return Plot.plot({
    width: 800,
    height: 290,
    x: { label: null },
    y: { label: "YoY change", tickFormat: d => `${d.toFixed(0)}%` },
    color: { domain: [true, false], range: ["#2ca02c", "#d62728"] },
    caption: "Green: growth vs. the same month a year earlier · Red: decline",
    marks: [
      Plot.ruleY([0], { stroke: "#999" }),
      Plot.rectY(data, {
        x: d => new Date(d.date),
        interval: "month",
        y: "yoy",
        fill: d => d.yoy >= 0,
        tip: { format: { fill: false, y: fmtSigned } }
      })
    ]
  })
}
```

# Snapshot

## Row

::: {.callout-note}
**Trailing 12-month window:** `{r} t12_label_cur` vs. `{r} t12_label_pri`. County growth is compared to the statewide rate (`{r} sprintf("%+.1f", state_t12_total_yoy)`%) over the same period. Counties without a full 12 months of data in both windows are excluded. Industry rankings reflect statewide totals only.
:::

## Row

```{ojs}
//| title: "Counties: Strongest & Weakest vs. State"

{
  const raw = transpose(county_t12_data)

  const N = 12  // top/bottom N to show

  const sorted = [...raw]
    .filter(d => d.yoy != null && !isNaN(d.yoy))
    .sort((a, b) => b.excess - a.excess)

  // Top N outperformers + bottom N underperformers
  const top    = sorted.slice(0, N)
  const bottom = sorted.slice(-N).reverse()
  const display = [
    ...top.map(d => ({ ...d, rank_group: "Outperforming" })),
    ...bottom.map(d => ({ ...d, rank_group: "Underperforming" }))
  ]

  // Sort for diverging bar: outperformers top, underperformers bottom
  display.sort((a, b) => b.excess - a.excess)

  return Plot.plot({
    width: 740,
    height: 560,
    marginLeft: 110,
    x: {
      label: "Excess growth vs. state T12 (%)",
      tickFormat: fmtSigned
    },
    y: { label: null },
    color: {
      domain: ["Outperforming", "Underperforming"],
      range: ["#2ca02c", "#d62728"],
      legend: true
    },
    caption: "Lighter bars: counties under $500M in trailing-12-month sales (small bases are more volatile)",
    marks: [
      Plot.barX(display, {
        x: "excess",
        y: "county",
        sort: { y: "-x" },
        fill: "rank_group",
        fillOpacity: d => d.small_county ? 0.5 : 0.85,
        tip: {
          format: {
            x: false,
            fill: false,
            fillOpacity: false
          },
          title: d =>
            `${d.county}${d.small_county ? " (small county)" : ""}\n` +
            `T12 growth: ${d.yoy != null ? fmtSigned(d.yoy) : "N/A"}\n` +
            `vs. state: ${d.excess != null ? fmtSigned(d.excess).replace("%", " ppts") : "N/A"}\n` +
            `T12 sales: ${fmtUSD(d.sales_current)}`
        }
      }),
      Plot.ruleX([0], { stroke: "#333", strokeWidth: 2 }),
      Plot.text(display.filter(d => d.excess >= 0), {
        x: "excess",
        y: "county",
        text: d => fmtSigned(d.excess),
        dx: 4,
        fontSize: 10,
        fill: "#2ca02c",
        textAnchor: "start"
      }),
      Plot.text(display.filter(d => d.excess < 0), {
        x: "excess",
        y: "county",
        text: d => fmtSigned(d.excess),
        dx: -4,
        fontSize: 10,
        fill: "#d62728",
        textAnchor: "end"
      })
    ]
  })
}
```

```{ojs}
//| title: "Industries: Statewide Strength & Weakness"

{
  const raw = transpose(state_ind_t12_data)
    .filter(d => d.state_yoy != null && !isNaN(d.state_yoy))
    .sort((a, b) => b.state_yoy - a.state_yoy)

  // Short display labels — full names in tooltip
  const display = raw.map(d => ({ ...d, label: shortInd(d.industry) }))

  return Plot.plot({
    width: 740,
    height: 420,
    marginLeft: 160,
    x: {
      label: "Trailing 12-month YoY growth (%)",
      tickFormat: fmtSigned
    },
    y: { label: null },
    color: {
      domain: [true, false],
      range: ["#2ca02c", "#d62728"],
      legend: true,
      tickFormat: d => d ? "Growing" : "Declining"
    },
    marks: [
      Plot.barX(display, {
        x: "state_yoy",
        y: "label",
        sort: { y: "-x" },
        fill: d => d.state_yoy >= 0,
        fillOpacity: 0.85,
        tip: {
          format: { x: false, fill: false, fillOpacity: false },
          title: d =>
            `${d.industry}\n` +
            `T12 growth: ${fmtSigned(d.state_yoy)}\n` +
            `T12 sales: ${fmtUSD(d.sales_t12)}`
        }
      }),
      Plot.ruleX([0], { stroke: "#333", strokeWidth: 2 }),
      Plot.text(display.filter(d => d.state_yoy >= 0), {
        x: "state_yoy",
        y: "label",
        text: d => fmtSigned(d.state_yoy),
        dx: 4,
        fontSize: 10,
        fill: "#2ca02c",
        textAnchor: "start"
      }),
      Plot.text(display.filter(d => d.state_yoy < 0), {
        x: "state_yoy",
        y: "label",
        text: d => fmtSigned(d.state_yoy),
        dx: -4,
        fontSize: 10,
        fill: "#d62728",
        textAnchor: "end"
      })
    ]
  })
}
```

# Geography

## Row {.sidebar}

```{ojs}
viewof geo_level = Inputs.radio(["County", "City"], { value: "County", label: "Geography:" })

viewof selected_geos = Inputs.select(
  geo_level === "County" ? counties_list : cities_list,
  {
    value: geo_level === "County" ? default_counties : default_cities,
    multiple: true,
    label: geo_level === "County" ? "Counties:" : "Cities:",
    size: 15
  }
)

viewof geo_metric = Inputs.radio(
  ["Retail Sales", "YoY Growth"],
  { value: "Retail Sales", label: "Metric:" }
)

viewof geo_time = Inputs.radio(["Monthly", "Annual"], { value: "Annual", label: "View:" })
```

::: {style="font-size: 0.8rem; color: #666; margin-top: 0.75rem;"}
Hold Ctrl (Cmd on Mac) to select multiple areas. Annual view includes complete calendar years only.
:::

## Row

```{ojs}
//| title: "Retail Sales Trends by Geography"

{
  if (selected_geos.length === 0) {
    return emptyNote("Select at least one area from the sidebar to see trends.")
  }

  const raw = geo_level === "County"
    ? transpose(county_data)
    : transpose(city_data)

  const geoCol = geo_level === "County" ? "county" : "city"
  const selected = new Set(selected_geos)
  const showYoy  = geo_metric === "YoY Growth"
  const isAnnual = geo_time === "Annual"

  let data = raw.filter(d => selected.has(d[geoCol]))

  if (isAnnual) {
    const grouped = d3.rollup(
      data.filter(d => completeYears.has(d.year)),
      v => ({
        [geoCol]: v[0][geoCol],
        year: v[0].year,
        retail_sales: d3.sum(v, d => d.retail_sales ?? 0)
      }),
      d => `${d[geoCol]}_${d.year}`
    )
    data = [...grouped.values()]
    // YoY must be computed after annual rollup
    if (showYoy) {
      const byGeo = d3.group(data, d => d[geoCol])
      data = [...byGeo.values()].flatMap(rows => {
        const sorted = rows.sort((a, b) => a.year - b.year)
        return sorted.map((d, i) => ({
          ...d,
          yoy: i === 0 ? null : (d.retail_sales / sorted[i-1].retail_sales - 1) * 100
        }))
      })
    }
  }

  const metric = showYoy ? "yoy" : "retail_sales"
  const yFmt   = showYoy ? (d => `${(+d).toFixed(1)}%`) : fmtUSD
  const yLabel = showYoy ? "YoY % change" : "Retail sales"

  // State YoY reference line — only meaningful in YoY mode (same scale as geos)
  let stateRef = []
  if (showYoy) {
    const stateRows = transpose(state_data)
    if (isAnnual) {
      const byYear = d3.rollup(
        stateRows.filter(d => completeYears.has(d.year)),
        v => d3.sum(v, d => d.retail_sales ?? 0),
        d => d.year
      )
      const years = [...byYear.keys()].sort()
      stateRef = years.slice(1).map(yr => ({
        year: yr,
        yoy: (byYear.get(yr) / byYear.get(yr - 1) - 1) * 100,
        label: "State"
      }))
    } else {
      stateRef = stateRows.filter(d => d.yoy != null && !isNaN(d.yoy)).map(d => ({...d, label: "State"}))
    }
  }

  return Plot.plot({
    width: 760,
    height: 360,
    color: { legend: true },
    x: isAnnual ? { label: null, tickFormat: d => String(d) } : { label: null },
    y: { label: yLabel, tickFormat: yFmt },
    marks: [
      Plot.lineY(data.filter(d => metric === "retail_sales" || (d[metric] != null && !isNaN(d[metric]))), {
        x: isAnnual ? "year" : (d => new Date(d.date)),
        y: metric,
        stroke: geoCol,
        tip: { format: { y: yFmt } },
        strokeWidth: 2
      }),
      // State YoY reference (only in YoY mode — same scale)
      ...(!showYoy ? [] : [
        Plot.lineY(stateRef, {
          x: isAnnual ? "year" : (d => new Date(d.date)),
          y: "yoy",
          stroke: "#888",
          strokeDasharray: "5,3",
          strokeWidth: 1.5
        }),
        Plot.text(stateRef.slice(-1), {
          x: isAnnual ? "year" : (d => new Date(d.date)),
          y: "yoy",
          text: () => "State avg",
          dy: -8,
          fontSize: 11,
          fill: "#888"
        })
      ])
    ]
  })
}
```

```{ojs}
//| title: "Top 20 by Retail Sales — Latest Month"

{
  const raw = geo_level === "County"
    ? transpose(county_data)
    : transpose(city_data)
  const geoCol = geo_level === "County" ? "county" : "city"

  // Most recent date
  const maxTs  = d3.max(raw, d => +new Date(d.date))
  const latest = raw.filter(d => +new Date(d.date) === maxTs && d.retail_sales != null)

  const stateRow = transpose(state_data).find(d => +new Date(d.date) === maxTs)
  const stateTotal = stateRow?.retail_sales

  const sorted = latest
    .sort((a, b) => b.retail_sales - a.retail_sales)
    .slice(0, 20)
    .map(d => ({
      ...d,
      share: stateTotal ? d.retail_sales / stateTotal * 100 : null
    }))

  const selected = new Set(selected_geos)

  return Plot.plot({
    width: 760,
    height: 480,
    marginLeft: 130,
    x: { label: "Retail sales", tickFormat: fmtUSD },
    y: { label: null },
    caption: `Highlighted bars are your sidebar selections · ${latest_month}`,
    marks: [
      Plot.barX(sorted, {
        x: "retail_sales",
        y: geoCol,
        sort: { y: "-x" },
        fill: d => selected.has(d[geoCol]) ? "#2E6DA4" : "#ccc",
        tip: {
          format: { x: false, fill: false },
          title: d =>
            `${d[geoCol]}\n` +
            `Retail sales: ${fmtUSD(d.retail_sales)}\n` +
            (d.share != null ? `Share of state: ${d.share.toFixed(1)}%` : "")
        }
      }),
      Plot.ruleX([0])
    ]
  })
}
```

# Industry

## Row {.sidebar}

```{ojs}
viewof ind_geo_level = Inputs.radio(["State", "County", "City"], { value: "State", label: "Geography:" })

viewof ind_selected_geo = {
  // Only offer geographies that actually have published industry detail
  const opts = ind_geo_level === "State"
    ? ["(Statewide)"]
    : ind_geo_level === "County" ? ind_counties_list : ind_cities_list
  return Inputs.select(opts, {
    value: opts.includes("Denver") ? "Denver" : opts[0],
    label: "Area:",
    disabled: ind_geo_level === "State"
  })
}

viewof ind_time = Inputs.radio(["Monthly", "Annual"], { value: "Annual", label: "View:" })

viewof selected_industries = Inputs.select(
  [...industries_active_list, ...industries_legacy_list],
  {
    value: default_industries,
    multiple: true,
    label: "Industries:",
    size: 14,
    format: d => legacySet.has(d) ? `${d} (thru 2021)` : d
  }
)
```

::: {style="font-size: 0.8rem; color: #666; margin-top: 0.75rem;"}
Industry detail is published only for large counties and cities. Categories marked "thru 2021" were retired in the 2022 NAICS revision; their successors begin in 2022.
:::

## Row

```{ojs}
//| title: "Retail Sales by Industry"

{
  let raw, geoCol

  if (ind_geo_level === "State") {
    raw = transpose(state_ind_data)
    geoCol = null
  } else if (ind_geo_level === "County") {
    raw = transpose(county_ind_data)
    geoCol = "county"
  } else {
    raw = transpose(city_ind_data)
    geoCol = "city"
  }

  let data = raw.filter(d =>
    (geoCol == null || d[geoCol] === ind_selected_geo) &&
    selected_industries.includes(d.industry) &&
    d.retail_sales != null
  )

  if (ind_time === "Annual") {
    const grouped = d3.rollup(
      data.filter(d => completeYears.has(d.year)),
      v => ({
        year: v[0].year,
        industry: v[0].industry,
        retail_sales: d3.sum(v, d => d.retail_sales ?? 0)
      }),
      d => `${d.industry}_${d.year}`
    )
    data = [...grouped.values()]
  }

  data = data.map(d => ({ ...d, ind_label: shortInd(d.industry) }))

  if (data.length === 0) {
    return emptyNote(
      "No industry data for this selection. Pick at least one industry — " +
      "county and city detail begins in 2022 for current categories."
    )
  }

  return Plot.plot({
    width: 760,
    height: 380,
    color: { legend: true },
    x: ind_time === "Annual" ? { label: null, tickFormat: d => String(d) } : { label: null },
    y: { label: "Retail sales", tickFormat: fmtUSD },
    marks: [
      Plot.lineY(data, {
        x: ind_time === "Annual" ? "year" : (d => new Date(d.date)),
        y: "retail_sales",
        stroke: "ind_label",
        tip: { format: { y: fmtUSD } },
        strokeWidth: 1.8
      })
    ]
  })
}
```

```{ojs}
//| title: "Industry Mix — Latest Month"

{
  let raw, geoCol

  if (ind_geo_level === "State") {
    raw = transpose(state_ind_data)
    geoCol = null
  } else if (ind_geo_level === "County") {
    raw = transpose(county_ind_data)
    geoCol = "county"
  } else {
    raw = transpose(city_ind_data)
    geoCol = "city"
  }

  const maxTs  = d3.max(raw, d => +new Date(d.date))
  const latest = raw.filter(d =>
    +new Date(d.date) === maxTs &&
    (geoCol == null || d[geoCol] === ind_selected_geo) &&
    d.industry !== "Total" &&
    d.retail_sales != null
  )

  if (latest.length === 0) {
    return emptyNote("No industry detail available for this area in the latest month.")
  }

  const total = d3.sum(latest, d => d.retail_sales)
  const sorted = [...latest]
    .sort((a, b) => b.retail_sales - a.retail_sales)
    .map(d => ({
      ...d,
      ind_label: shortInd(d.industry),
      share: total > 0 ? d.retail_sales / total * 100 : null
    }))

  return Plot.plot({
    width: 760,
    height: 460,
    marginLeft: 175,
    x: { label: "Retail sales", tickFormat: fmtUSD },
    y: { label: null },
    marks: [
      Plot.barX(sorted, {
        x: "retail_sales",
        y: "ind_label",
        sort: { y: "-x" },
        fill: "#2E6DA4",
        fillOpacity: 0.8,
        tip: {
          format: { x: false, fill: false },
          title: d =>
            `${d.industry}\n` +
            `Retail sales: ${fmtUSD(d.retail_sales)}\n` +
            (d.share != null ? `Share of area total: ${d.share.toFixed(1)}%` : "")
        }
      }),
      Plot.ruleX([0])
    ]
  })
}
```

# Area × Industry

## Row {.sidebar}

```{ojs}
viewof hm_metric = Inputs.radio(
  ["Excess vs. state industry", "Raw YoY growth"],
  { value: "Excess vs. state industry", label: "Color metric:" }
)

viewof hm_clamp = Inputs.range(
  [5, 40],
  { value: 20, step: 5, label: "Color scale cap (±%):" }
)
```

::: {.callout-note style="font-size: 0.85rem; margin-top: 1rem;"}
**Coverage:** Industry detail is available for the **`{r} length(ind_counties)` major counties** (`{r} paste(ind_counties, collapse = ", ")`) for 2022–present. Color scale is capped at the selected cap value; actual values appear on hover.`{r} weld_note`
:::

## Row

```{ojs}
//| title: "County × Industry Growth Heatmap (Trailing 12 Months)"

{
  const raw = transpose(heatmap_data)

  const metricCol = hm_metric === "Excess vs. state industry" ? "excess" : "yoy"
  const cap       = hm_clamp  // user-controlled color scale cap

  const data = raw
    .filter(d => d[metricCol] != null && !isNaN(d[metricCol]))
    .map(d => ({
      ...d,
      ind_label:     shortInd(d.industry),
      value:         d[metricCol],
      value_clamped: Math.max(-cap, Math.min(cap, d[metricCol]))
    }))

  // Sort industry y-axis by statewide T12 YoY (strongest at top)
  const indOrder = [...d3.rollup(data, v => v[0].state_yoy, d => d.ind_label)]
    .sort((a, b) => b[1] - a[1])
    .map(d => d[0])

  const colorLabel = hm_metric === "Excess vs. state industry"
    ? `Excess growth vs. state industry (ppts, capped ±${cap}%)`
    : `YoY growth (%, capped ±${cap}%)`

  return Plot.plot({
    width:      820,
    height:     420,
    marginLeft: 165,
    marginBottom: 80,
    color: {
      type:   "diverging",
      scheme: "RdYlGn",
      domain: [-cap, cap],
      label:  colorLabel,
      legend: true
    },
    x: {
      label:      null,
      tickRotate: -35,
      domain:     [...new Set(data.map(d => d.county))].sort()
    },
    y: {
      label:  null,
      domain: indOrder
    },
    caption: `Trailing 12 months: ${t12_label_cur} · Outlined cells exceed the color cap`,
    marks: [
      Plot.cell(data, {
        x:    "county",
        y:    "ind_label",
        fill: "value_clamped",
        tip: {
          format: { fill: false },
          title: d => {
            const metricLabel = hm_metric === "Excess vs. state industry"
              ? `Excess vs. state: ${fmtSigned(d.excess).replace("%", " ppts")}`
              : `YoY growth: ${fmtSigned(d.yoy)}`
            const stateRef = `State industry T12: ${fmtSigned(d.state_yoy)}`
            return `${d.county}\n${d.industry}\n${metricLabel}\n${stateRef}`
          }
        }
      }),
      // Outline cells that are beyond the color cap (extreme values)
      Plot.cell(data.filter(d => Math.abs(d.value) > cap), {
        x:             "county",
        y:             "ind_label",
        stroke:        "#333",
        strokeWidth:   1.5,
        fill:          "none"
      })
    ]
  })
}
```

```{ojs}
//| title: "Notable Area × Industry Combinations"

{
  const raw = transpose(heatmap_data)
    .filter(d => d.excess != null && !isNaN(d.excess) && d.yoy != null && !isNaN(d.yoy))

  const metricCol = hm_metric === "Excess vs. state industry" ? "excess" : "yoy"

  // Top 8 + bottom 8 county×industry combos by chosen metric
  const sorted = [...raw].sort((a, b) => b[metricCol] - a[metricCol])
  const top    = sorted.slice(0, 8).map(d => ({ ...d, group: "Standout" }))
  const bottom = sorted.slice(-8).reverse().map(d => ({ ...d, group: "Lagging" }))
  const display = [...top, ...bottom]
    .sort((a, b) => b[metricCol] - a[metricCol])
    .map(d => ({
      ...d,
      label: `${d.county} / ${shortInd(d.industry)}`
    }))

  const xLabel = hm_metric === "Excess vs. state industry"
    ? "Excess growth vs. state industry (ppts)"
    : "YoY growth (%)"

  return Plot.plot({
    width:       740,
    height:      520,
    marginLeft:  230,
    x: {
      label:      xLabel,
      tickFormat: fmtSigned
    },
    y: { label: null },
    color: {
      domain: ["Standout", "Lagging"],
      range:  ["#2ca02c", "#d62728"],
      legend: true
    },
    marks: [
      Plot.dot(display, {
        x:    metricCol,
        y:    "label",
        sort: { y: "-x" },
        r:    6,
        fill: "group",
        tip: {
          format: { fill: false, r: false },
          title: d =>
            `${d.county}\n${d.industry}\n` +
            `T12 YoY: ${fmtSigned(d.yoy)}\n` +
            `State industry T12: ${fmtSigned(d.state_yoy)}\n` +
            `Excess: ${fmtSigned(d.excess).replace("%", " ppts")}`
        }
      }),
      Plot.ruleX([0], { stroke: "#333", strokeWidth: 1.5 }),
      Plot.text(display.filter(d => d[metricCol] >= 0), {
        x:          metricCol,
        y:          "label",
        text:       d => fmtSigned(d[metricCol]),
        dx:         10,
        fontSize:   10,
        fill:       "#2ca02c",
        textAnchor: "start"
      }),
      Plot.text(display.filter(d => d[metricCol] < 0), {
        x:          metricCol,
        y:          "label",
        text:       d => fmtSigned(d[metricCol]),
        dx:         -10,
        fontSize:   10,
        fill:       "#d62728",
        textAnchor: "end"
      })
    ]
  })
}
```

# Annual Performance

## Row

::: {.callout-note}
**Excess growth** compares each area's annual retail sales growth to the statewide rate for the same year (and industry, if one is selected). Only complete calendar years are offered, and each area must have a full 12 months of data in both years. Industry-level comparisons cover only the large counties and cities with published industry detail; categories introduced in the 2022 NAICS revision support 2023 onward.
:::

```{ojs}
//| title: "Controls"

viewof out_geo_level = Inputs.radio(["County", "City"], { value: "County", label: "Geography:" })

viewof out_year = Inputs.select(
  complete_years.filter(y => complete_years.includes(y - 1)).sort((a, b) => b - a),
  { label: "Year:" }
)

viewof out_industry = Inputs.select(
  ["All Industries", ...industries_active_list],
  { value: "All Industries", label: "Industry:" }
)
```

## Row

```{ojs}
//| title: "Excess Growth vs. State — Top & Bottom Performers"

{
  const useInd = out_industry !== "All Industries"
  const geoCol = out_geo_level === "County" ? "county" : "city"

  const raw = useInd
    ? (out_geo_level === "County" ? transpose(county_ind_data) : transpose(city_ind_data))
        .filter(d => d.industry === out_industry)
    : (out_geo_level === "County" ? transpose(county_data) : transpose(city_data))

  // Statewide benchmark (industry-specific when one is selected)
  const stateRows = useInd
    ? transpose(state_ind_data).filter(d => d.industry === out_industry)
    : transpose(state_data)

  const stateByYear = d3.rollup(
    stateRows.filter(d => d.retail_sales != null),
    v => d3.sum(v, d => d.retail_sales),
    d => d.year
  )
  const sCur  = stateByYear.get(out_year)
  const sPrev = stateByYear.get(out_year - 1)
  if (!sCur || !sPrev) {
    return emptyNote(
      `No statewide data for ${out_industry} in ${out_year - 1}–${out_year}. ` +
      "Categories introduced in the 2022 NAICS revision support 2023 onward."
    )
  }
  const stateGrowth = (sCur / sPrev - 1) * 100

  // Annual totals per geography, with month counts to screen out gaps
  const geoYear = d3.rollup(
    raw.filter(d => d.retail_sales != null),
    v => ({
      sales: d3.sum(v, d => d.retail_sales),
      months: new Set(v.map(d => d.month)).size
    }),
    d => d[geoCol],
    d => d.year
  )

  const results = [...geoYear.entries()].flatMap(([geo, byYear]) => {
    const cur  = byYear.get(out_year)
    const prev = byYear.get(out_year - 1)
    if (!cur || !prev) return []
    if (cur.months < 12 || prev.months < 12) return []
    if (prev.sales <= 0 || cur.sales <= 0) return []
    const growth = (cur.sales / prev.sales - 1) * 100
    return [{ geo, growth, excess: growth - stateGrowth, sales: cur.sales }]
  })

  if (results.length === 0) {
    return emptyNote("No areas have a full 12 months of data in both years for this selection.")
  }

  const N = 12
  const sorted  = [...results].sort((a, b) => b.excess - a.excess)
  const display = sorted.length <= 2 * N
    ? sorted
    : [...sorted.slice(0, N), ...sorted.slice(-N)]

  return Plot.plot({
    width: 780,
    height: 540,
    marginLeft: 140,
    x: {
      label: "Excess growth vs. state (%)",
      tickFormat: fmtSigned
    },
    y: { label: null },
    color: {
      domain: [true, false],
      range: ["#2ca02c", "#d62728"],
      legend: true,
      tickFormat: d => d ? "Outperforming" : "Underperforming"
    },
    caption: "Areas with a full 12 months of data in both years",
    marks: [
      Plot.barX(display, {
        x: "excess",
        y: "geo",
        sort: { y: "-x" },
        fill: d => d.excess >= 0,
        tip: {
          format: { x: false, fill: false },
          title: d =>
            `${d.geo}\n` +
            `${out_year} growth: ${fmtSigned(d.growth)}\n` +
            `State: ${fmtSigned(stateGrowth)}\n` +
            `Excess: ${fmtSigned(d.excess).replace("%", " ppts")}\n` +
            `${out_year} sales: ${fmtUSD(d.sales)}`
        }
      }),
      Plot.ruleX([0], { stroke: "#333", strokeWidth: 2 }),
      Plot.text(display.filter(d => d.excess >= 0), {
        x: "excess",
        y: "geo",
        text: d => fmtSigned(d.excess),
        dx: 4,
        fontSize: 11,
        fill: "#2ca02c",
        textAnchor: "start"
      }),
      Plot.text(display.filter(d => d.excess < 0), {
        x: "excess",
        y: "geo",
        text: d => fmtSigned(d.excess),
        dx: -4,
        fontSize: 11,
        fill: "#d62728",
        textAnchor: "end"
      })
    ]
  })
}
```

```{ojs}
//| title: "State Benchmark"

{
  const useInd = out_industry !== "All Industries"
  const stateRows = useInd
    ? transpose(state_ind_data).filter(d => d.industry === out_industry)
    : transpose(state_data)

  const stateByYear = d3.rollup(
    stateRows.filter(d => d.retail_sales != null),
    v => d3.sum(v, d => d.retail_sales),
    d => d.year
  )
  const sCur  = stateByYear.get(out_year)
  const sPrev = stateByYear.get(out_year - 1)

  if (!sCur || !sPrev) {
    return emptyNote("No statewide benchmark available for this selection.")
  }

  const stateGrowth = (sCur / sPrev - 1) * 100
  const label = useInd ? `${out_industry} —` : ""

  return html`<div style="padding: 1rem; background: #f0f4ff; border-radius: 8px; font-size: 1.1rem;">
    <strong>${label} State YoY (${out_year}):</strong>
    <span style="color: ${stateGrowth >= 0 ? '#2ca02c' : '#d62728'}; font-weight: bold;">
      ${fmtSigned(stateGrowth)}
    </span>
    <br/>Areas above this rate are outperforming the state average.
  </div>`
}
```

# About

## Row

::: {.card title="Data source & definitions"}
**Source:** [Colorado Department of Revenue retail sales reports](https://cdor.colorado.gov/retail-sales-reports), based on state sales tax returns.

**Coverage:** Statewide monthly data from January 2016 through `{r} format(latest_date, "%B %Y")`; all 64 counties; ~`{r} length(cities)` cities. Industry detail is published only for `{r} length(ind_counties)` large counties (`{r} paste(ind_counties, collapse = ", ")`) and `{r} length(ind_cities)` large cities.

**Definitions:**

- **Year-over-year (YoY) growth** — change vs. the same month (or year) one year earlier.
- **Trailing 12 months (T12)** — the most recent 12 months of data (`{r} t12_label_cur`), compared with the preceding 12 months (`{r} t12_label_pri`). T12 smooths out month-to-month noise and seasonality.
- **Excess growth** — an area's or industry's growth rate minus the comparable statewide rate, in percentage points. Positive values mean faster-than-state growth.

**Refreshing the data:** run `R/collect_data.R` to download the latest reports and rebuild the parquet files, then re-render this dashboard. The original Excel workbooks are saved to `data/raw/` for manual inspection.

**Exporting charts:** every chart card has buttons in its header — copy puts a high-resolution PNG on the clipboard; download saves a scalable SVG that can be resized freely in presentations. (Copy requires a regular browser; embedded viewers such as IDE preview panes may block clipboard access.)
:::

::: {.card title="Caveats"}
- Figures are **not seasonally adjusted** and **not inflation-adjusted**. Use YoY or T12 comparisons rather than month-to-month changes.
- The **2022 NAICS revision** redefined industry categories. Retired categories are marked "(thru 2021)" in the Industry tab; growth comparisons spanning 2021–2022 partly reflect reclassification rather than real change.
- Small counties and cities have volatile growth rates; the Snapshot tab de-emphasizes counties under $500M in annual sales.
- Some city and county values are suppressed or revised by the Department of Revenue; annual and trailing-12-month comparisons in this dashboard require a full 12 months of data in both periods.
- Known anomaly: Weld County / Sporting Goods shows an unexplained spike in December 2025.
:::

```{=html}
<script>
(function() {
  // Adds copy/download buttons to every dashboard card that contains a chart,
  // so charts can be dropped straight into presentations. Copy puts a PNG on
  // the clipboard (clipboards don't accept SVG); download saves a scalable
  // SVG built from native SVG elements only — no foreignObject — so the file
  // opens correctly in PowerPoint and other Office apps.

  function cardFileName(card, ext) {
    const header = card.querySelector(".card-header")
    const text = header ? header.textContent.trim() : "chart"
    return (text.replace(/[^\w\- ]+/g, "").replace(/\s+/g, "_").toLowerCase() || "chart") + "." + ext
  }

  const SVG_NS = "http://www.w3.org/2000/svg"
  const CAPTION_FONT = "11px system-ui, sans-serif"
  const LEGEND_FONT = "11px system-ui, sans-serif"

  function measureText(text, font) {
    const ctx = measureText._ctx ?? (measureText._ctx = document.createElement("canvas").getContext("2d"))
    ctx.font = font
    return ctx.measureText(text).width
  }

  function wrapText(text, font, maxWidth) {
    const lines = []
    let line = ""
    for (const word of text.split(/\s+/)) {
      const candidate = line ? line + " " + word : word
      if (line && measureText(candidate, font) > maxWidth) {
        lines.push(line)
        line = word
      } else {
        line = candidate
      }
    }
    if (line) lines.push(line)
    return lines
  }

  function svgText(content, x, y, font, fill) {
    const t = document.createElementNS(SVG_NS, "text")
    t.setAttribute("x", x)
    t.setAttribute("y", y)
    t.setAttribute("style", `font: ${font}; fill: ${fill};`)
    t.textContent = content
    return t
  }

  // Rebuilds the Plot figure as one standalone SVG: nested chart svg(s) kept
  // as-is, HTML swatch legends redrawn as rects + text, figcaption as text.
  function figureToSvgEl(fig) {
    const pad = 8
    const parts = []  // { el, height } stacked vertically
    let width = 0

    for (const child of fig.children) {
      if (child.tagName === "SVG" || child instanceof SVGSVGElement) {
        // Chart itself, or an svg-based ramp legend — embed directly
        const clone = child.cloneNode(true)
        clone.style.maxWidth = "none"
        const w = +child.getAttribute("width") || child.getBoundingClientRect().width
        const h = +child.getAttribute("height") || child.getBoundingClientRect().height
        width = Math.max(width, w)
        parts.push({ el: clone, height: h, isSvg: true })
      } else if (child.tagName === "DIV") {
        // HTML swatch legend: one row of colored squares with labels
        const swatches = [...child.querySelectorAll("span")].filter(s => s.querySelector("svg"))
        if (swatches.length === 0) continue
        const g = document.createElementNS(SVG_NS, "g")
        let x = pad
        for (const sw of swatches) {
          const color = sw.querySelector("svg")?.getAttribute("fill") || "#888"
          const label = sw.textContent.trim()
          const rect = document.createElementNS(SVG_NS, "rect")
          rect.setAttribute("x", x)
          rect.setAttribute("y", 2)
          rect.setAttribute("width", 12)
          rect.setAttribute("height", 12)
          rect.setAttribute("fill", color)
          g.appendChild(rect)
          g.appendChild(svgText(label, x + 16, 12, LEGEND_FONT, "#333"))
          x += 16 + measureText(label, LEGEND_FONT) + 14
        }
        width = Math.max(width, x)
        parts.push({ el: g, height: 20 })
      } else if (child.tagName === "FIGCAPTION") {
        const g = document.createElementNS(SVG_NS, "g")
        const lines = wrapText(child.textContent.trim(), CAPTION_FONT, Math.max(width - 2 * pad, 200))
        lines.forEach((line, i) => g.appendChild(svgText(line, pad, 11 + i * 15, CAPTION_FONT, "#666")))
        parts.push({ el: g, height: lines.length * 15 + 4 })
      }
    }

    if (!parts.some(p => p.isSvg)) throw new Error("No chart in this card")

    const totalHeight = parts.reduce((sum, p) => sum + p.height, 0) + pad
    const out = document.createElementNS(SVG_NS, "svg")
    out.setAttribute("width", width)
    out.setAttribute("height", totalHeight)
    out.setAttribute("viewBox", `0 0 ${width} ${totalHeight}`)

    const bg = document.createElementNS(SVG_NS, "rect")
    bg.setAttribute("width", width)
    bg.setAttribute("height", totalHeight)
    bg.setAttribute("fill", "#fff")
    out.appendChild(bg)

    let y = pad / 2
    for (const part of parts) {
      if (part.isSvg) {
        part.el.setAttribute("y", y)
      } else {
        part.el.setAttribute("transform", `translate(0, ${y})`)
      }
      out.appendChild(part.el)
      y += part.height
    }
    return out
  }

  function figureToSvgBlob(fig) {
    const xml = '<?xml version="1.0" encoding="UTF-8"?>\n' +
      new XMLSerializer().serializeToString(figureToSvgEl(fig))
    return new Blob([xml], { type: "image/svg+xml;charset=utf-8" })
  }

  async function figureToPngBlob(fig) {
    // Clone and measure at natural (unscaled) size off-screen; the page CSS
    // shrinks wide charts to fit their cards, but exports should be full size.
    const clone = fig.cloneNode(true)
    clone.querySelectorAll("svg").forEach(s => { s.style.maxWidth = "none" })
    const meas = document.createElement("div")
    meas.style.cssText = "position:fixed;left:-10000px;top:0;width:max-content;background:#fff;"
    meas.appendChild(clone)
    document.body.appendChild(meas)
    const rect = meas.getBoundingClientRect()
    const w = Math.ceil(rect.width)
    const h = Math.ceil(rect.height)
    const xml = new XMLSerializer().serializeToString(clone)
    meas.remove()

    const svg =
      `<svg xmlns="http://www.w3.org/2000/svg" width="${w}" height="${h}">` +
      `<foreignObject width="100%" height="100%">` +
      `<div xmlns="http://www.w3.org/1999/xhtml" style="margin:0;background:#fff;` +
      `font-family:system-ui,-apple-system,'Segoe UI',sans-serif;">` +
      `<style>figure{margin:0;padding:6px;}figcaption{font-size:11px;color:#666;padding-top:4px;}</style>` +
      xml +
      `</div></foreignObject></svg>`

    const img = new Image()
    await new Promise((resolve, reject) => {
      img.onload = resolve
      img.onerror = () => reject(new Error("Could not rasterize chart"))
      img.src = "data:image/svg+xml;charset=utf-8," + encodeURIComponent(svg)
    })

    const scale = 2  // 2x resolution for crisp slides
    const canvas = document.createElement("canvas")
    canvas.width = w * scale
    canvas.height = h * scale
    const ctx = canvas.getContext("2d")
    ctx.fillStyle = "#fff"
    ctx.fillRect(0, 0, canvas.width, canvas.height)
    ctx.drawImage(img, 0, 0, canvas.width, canvas.height)
    return new Promise((resolve, reject) =>
      canvas.toBlob(b => b ? resolve(b) : reject(new Error("PNG export failed")), "image/png")
    )
  }

  function flash(btn, ok) {
    const original = btn.innerHTML
    btn.innerHTML = ok ? '<i class="bi bi-check-lg"></i>' : '<i class="bi bi-x-lg"></i>'
    btn.disabled = true
    setTimeout(() => { btn.innerHTML = original; btn.disabled = false }, 1500)
  }

  function makeButton(iconClass, titleText, onClick) {
    const btn = document.createElement("button")
    btn.type = "button"
    btn.className = "btn btn-sm btn-outline-secondary chart-export-btn"
    btn.title = titleText
    btn.setAttribute("aria-label", titleText)
    btn.innerHTML = `<i class="bi ${iconClass}"></i>`
    btn.style.cssText = "padding:0 .45rem;font-size:.75rem;line-height:1.5;"
    btn.addEventListener("click", async (ev) => {
      ev.preventDefault()
      ev.stopPropagation()
      try {
        await onClick()
        flash(btn, true)
      } catch (err) {
        console.error("Chart export failed:", err)
        flash(btn, false)
      }
    })
    return btn
  }

  function addButtons(card) {
    const header = card.querySelector(".card-header")
    if (!header || header.querySelector(".chart-export-btn")) return
    // Re-query at click time: OJS replaces the figure when inputs change
    const getFigure = () => card.querySelector(".card-body figure")

    const group = document.createElement("div")
    group.className = "chart-export-group"
    group.style.cssText = "display:inline-flex;gap:.25rem;margin-left:auto;"

    group.appendChild(makeButton("bi-copy", "Copy chart to clipboard (PNG)", async () => {
      const fig = getFigure()
      if (!fig) throw new Error("No chart in this card")
      // Pass the pending blob promise straight to ClipboardItem — awaiting it
      // first would expire the user-gesture window in Safari
      await navigator.clipboard.write([
        new ClipboardItem({ "image/png": figureToPngBlob(fig) })
      ])
    }))

    group.appendChild(makeButton("bi-download", "Download chart as SVG", async () => {
      const fig = getFigure()
      if (!fig) throw new Error("No chart in this card")
      const blob = figureToSvgBlob(fig)
      const a = document.createElement("a")
      a.href = URL.createObjectURL(blob)
      a.download = cardFileName(card, "svg")
      a.click()
      setTimeout(() => URL.revokeObjectURL(a.href), 5000)
    }))

    header.style.display = "flex"
    header.style.alignItems = "center"
    header.appendChild(group)
  }

  function scanCards() {
    document.querySelectorAll(".card").forEach(card => {
      if (card.querySelector(".card-body figure svg")) addButtons(card)
    })
  }

  // OJS renders asynchronously (and re-renders on input changes), so watch
  // the DOM rather than polling — timers are throttled in background tabs
  scanCards()
  let scanScheduled = false
  new MutationObserver(() => {
    if (scanScheduled) return
    scanScheduled = true
    requestAnimationFrame(() => {
      scanScheduled = false
      scanCards()
    })
  }).observe(document.body, { childList: true, subtree: true })
})()
</script>
```
