---
title: "Colorado Retail Sales"
format:
  dashboard:
    theme: cosmo
    embed-resources: true
    nav-buttons:
      - icon: github
        href: https://cdor.colorado.gov/retail-sales-reports
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
library(tidyverse)
library(arrow)
library(scales)
if (requireNamespace("sdotools", quietly = TRUE)) library(sdotools)

# Load pre-built parquet files (run R/collect_data.R to refresh)
state_df          <- read_parquet("data/state.parquet")
county_df         <- read_parquet("data/county.parquet")
city_df           <- read_parquet("data/city.parquet")
county_ind_df     <- read_parquet("data/county_industry.parquet")
city_ind_df       <- read_parquet("data/city_industry.parquet")
state_ind_df      <- read_parquet("data/state_industry.parquet")

# Helpers -----------------------------------------------------------------

yoy_growth <- function(df, value_col = retail_sales, group_cols = c()) {
  df |>
    arrange(across(all_of(c(group_cols, "year", "month")))) |>
    group_by(across(all_of(group_cols))) |>
    mutate(
      yoy = (.data[[deparse(substitute(value_col))]] /
               lag(.data[[deparse(substitute(value_col))]], 12) - 1) * 100
    ) |>
    ungroup()
}

# Compute YoY for state
state_yoy <- state_df |>
  arrange(year, month) |>
  mutate(yoy = (retail_sales / lag(retail_sales, 12) - 1) * 100)

# Latest month for headlines
latest <- state_df |> slice_max(date, n = 1)
prev_yr <- state_df |>
  filter(year == latest$year - 1L, month == latest$month) |>
  slice(1)

# Counties list (excluding "Total" aggregate)
counties <- county_df |>
  filter(county != "Total") |>
  distinct(county) |>
  pull(county) |>
  sort()

cities <- city_df |>
  distinct(city) |>
  pull(city) |>
  sort()

industries <- state_ind_df |>
  filter(industry != "Total") |>
  distinct(industry) |>
  pull(industry) |>
  sort()

# ---------------------------------------------------------------------------
# Trailing-12-month (T12) precomputation for Snapshot and Heatmap tabs
# T12 window: Mar 2025 – Feb 2026  vs.  Mar 2024 – Feb 2025
# ---------------------------------------------------------------------------

t12_start_cur  <- as.Date("2025-03-01")
t12_end_cur    <- as.Date("2026-02-01")
t12_start_pri  <- as.Date("2024-03-01")
t12_end_pri    <- as.Date("2025-02-01")

assign_period <- function(df) {
  df |>
    mutate(period = case_when(
      date >= t12_start_cur & date <= t12_end_cur ~ "current",
      date >= t12_start_pri & date <= t12_end_pri ~ "prior",
      TRUE ~ NA_character_
    ))
}

# State T12 total (for county excess benchmark)
state_t12_total_yoy <- county_df |>
  filter(county == "Total", !is.na(retail_sales)) |>
  assign_period() |>
  filter(!is.na(period)) |>
  group_by(period) |>
  summarise(sales = sum(retail_sales, na.rm = TRUE), .groups = "drop") |>
  pivot_wider(names_from = period, values_from = sales) |>
  mutate(state_yoy = (current / prior - 1) * 100) |>
  pull(state_yoy)

# County T12 YoY — all 64 counties
county_t12 <- county_df |>
  filter(county != "Total", !is.na(retail_sales)) |>
  assign_period() |>
  filter(!is.na(period)) |>
  group_by(county, period) |>
  summarise(
    sales     = sum(retail_sales, na.rm = TRUE),
    n_months  = n(),
    .groups   = "drop"
  ) |>
  pivot_wider(names_from = period, values_from = c(sales, n_months)) |>
  mutate(
    yoy    = (sales_current / sales_prior - 1) * 100,
    excess = yoy - state_t12_total_yoy,
    # Flag small counties (prior-period T12 < $500M) for context
    small_county = sales_prior < 500e6
  )

# Core retail industries (post-2022 NAICS, 0% NA rate in county_industry)
core_heatmap_inds <- c(
  "Accommodation",
  "Building Material and Garden Equipment and Supplies Dealers",
  "Clothing, Clothing Accessories, Shoe, and Jewelry Retailers",
  "Food and Beverage Retailers",
  "Food Services and Drinking Places",
  "Furniture, Home Furnishings, Electronics, and Appliance Retailers",
  "Gasoline Stations and Fuel Dealers",
  "General Merchandise Retailers",
  "Health and Personal Care Retailers",
  "Motor Vehicle and Parts Dealers",
  "Sporting Goods, Hobby, Musical Instrument, Book, and Miscellaneous Retailers"
)

# State T12 YoY by industry (benchmark for heatmap excess)
state_ind_t12 <- state_ind_df |>
  filter(industry %in% core_heatmap_inds, !is.na(retail_sales)) |>
  assign_period() |>
  filter(!is.na(period)) |>
  group_by(industry, period) |>
  summarise(sales = sum(retail_sales, na.rm = TRUE), .groups = "drop") |>
  pivot_wider(names_from = period, values_from = sales) |>
  mutate(
    state_yoy   = (current / prior - 1) * 100,
    sales_t12   = current
  )

# County × industry T12 YoY with excess vs state industry benchmark
heatmap_county_ind <- county_ind_df |>
  filter(industry %in% core_heatmap_inds, !is.na(retail_sales)) |>
  assign_period() |>
  filter(!is.na(period)) |>
  group_by(county, industry, period) |>
  summarise(sales = sum(retail_sales, na.rm = TRUE), .groups = "drop") |>
  pivot_wider(names_from = period, values_from = sales) |>
  mutate(yoy = (current / prior - 1) * 100) |>
  left_join(
    state_ind_t12 |> select(industry, state_yoy),
    by = "industry"
  ) |>
  mutate(excess = yoy - state_yoy)

# Pass key data to OJS
ojs_define(
  state_data          = state_yoy,
  county_data         = county_df |>
    filter(county != "Total") |>
    arrange(county, year, month) |>
    group_by(county) |>
    mutate(yoy = (retail_sales / lag(retail_sales, 12) - 1) * 100) |>
    ungroup(),
  city_data           = city_df |>
    arrange(city, year, month) |>
    group_by(city) |>
    mutate(yoy = (retail_sales / lag(retail_sales, 12) - 1) * 100) |>
    ungroup(),
  county_ind_data     = county_ind_df,
  city_ind_data       = city_ind_df,
  state_ind_data      = state_ind_df,
  counties_list       = counties,
  cities_list         = cities,
  industries_list     = industries,
  latest_month        = format(latest$date, "%B %Y"),
  latest_sales        = latest$retail_sales,
  latest_yoy          = round(state_yoy |> slice_max(date, n = 1) |> pull(yoy), 1),
  # New: pre-aggregated T12 data for Snapshot and Heatmap tabs
  county_t12_data     = county_t12,
  state_ind_t12_data  = state_ind_t12,
  heatmap_data        = heatmap_county_ind,
  state_t12_yoy       = round(state_t12_total_yoy, 2)
)
```

# State Overview

## Row {height="15%"}

```{r}
#| content: valuebox
#| title: "Statewide Retail Sales"
list(
  icon  = "shop",
  color = "primary",
  value = dollar(latest$retail_sales, scale = 1e-9, suffix = "B", accuracy = 0.1)
)
```

```{r}
#| content: valuebox
#| title: "Year-over-Year Change"
yoy_val <- state_yoy |> slice_max(date, n = 1) |> pull(yoy)
list(
  icon  = if (yoy_val >= 0) "arrow-up" else "arrow-down",
  color = if (yoy_val >= 0) "success" else "danger",
  value = paste0(sprintf("%+.1f", yoy_val), "%")
)
```

```{r}
#| content: valuebox
#| title: "Latest Month"
list(
  icon  = "calendar3",
  color = "secondary",
  value = format(latest$date, "%B %Y")
)
```

## Row

```{ojs}
//| title: "Monthly Retail Sales (Statewide)"

import { Plot } from "@observablehq/plot"

viewof time_unit = Inputs.radio(["Monthly", "Annual"], { value: "Monthly", label: "View:" })

{
  const data = transpose(state_data)
  let plotData = data.filter(d => !isNaN(d.retail_sales))

  if (time_unit === "Annual") {
    const byYear = d3.rollup(plotData, v => ({
      year: v[0].year,
      retail_sales: d3.sum(v, d => d.retail_sales)
    }), d => d.year)
    plotData = [...byYear.values()]
  }

  return Plot.plot({
    width: 800,
    height: 320,
    x: { label: null, tickFormat: d => time_unit === "Annual" ? String(d) : undefined },
    y: { label: "Retail Sales ($)", tickFormat: d => `$${(d/1e9).toFixed(1)}B` },
    marks: [
      Plot.areaY(plotData, {
        x: time_unit === "Annual" ? "year" : (d => new Date(d.date)),
        y: "retail_sales",
        fill: "#2E6DA4",
        fillOpacity: 0.2
      }),
      Plot.lineY(plotData, {
        x: time_unit === "Annual" ? "year" : (d => new Date(d.date)),
        y: "retail_sales",
        stroke: "#2E6DA4",
        strokeWidth: 2,
        tip: true
      })
    ]
  })
}
```

```{ojs}
//| title: "Year-over-Year Growth Rate"

{
  const data = transpose(state_data).filter(d => d.yoy != null && !isNaN(d.yoy))
  return Plot.plot({
    width: 800,
    height: 260,
    x: { label: null },
    y: { label: "YoY %", tickFormat: d => `${d.toFixed(1)}%` },
    marks: [
      Plot.ruleY([0], { stroke: "#999" }),
      Plot.lineY(data, {
        x: d => new Date(d.date),
        y: "yoy",
        stroke: d => d.yoy >= 0 ? "#2ca02c" : "#d62728",
        strokeWidth: 2,
        tip: true
      })
    ]
  })
}
```

# Geography

## Row {.sidebar}

```{ojs}
viewof geo_level = Inputs.radio(["County", "City"], { value: "County", label: "Geography:" })

viewof selected_geos = Inputs.select(
  geo_level === "County" ? counties_list : cities_list,
  {
    value: geo_level === "County"
      ? counties_list.slice(0, 5)
      : cities_list.filter(c => ["Denver", "Aurora", "Colorado Springs", "Fort Collins", "Boulder"].includes(c)),
    multiple: true,
    label: geo_level === "County" ? "Counties:" : "Cities:",
    size: 15
  }
)

viewof geo_metric = Inputs.radio(
  ["Retail Sales", "YoY Growth"],
  { value: "Retail Sales", label: "Metric:" }
)

viewof geo_time = Inputs.radio(["Monthly", "Annual"], { value: "Annual", label: "View:" })
```

## Row

```{ojs}
//| title: "Retail Sales by Geography vs. State Average"

{
  const raw = geo_level === "County"
    ? transpose(county_data)
    : transpose(city_data)

  const geoCol = geo_level === "County" ? "county" : "city"
  const selected = new Set(selected_geos)
  const showYoy  = geo_metric === "YoY Growth"
  const isAnnual = geo_time === "Annual"

  let data = raw.filter(d => selected.has(d[geoCol]))

  if (isAnnual) {
    const grouped = d3.rollup(data, v => ({
      [geoCol]: v[0][geoCol],
      year: v[0].year,
      retail_sales: d3.sum(v, d => d.retail_sales ?? 0)
    }), d => `${d[geoCol]}_${d.year}`)
    data = [...grouped.values()]
    // YoY must be computed after annual rollup
    if (showYoy) {
      const byGeo = d3.group(data, d => d[geoCol])
      data = [...byGeo.values()].flatMap(rows => {
        const sorted = rows.sort((a, b) => a.year - b.year)
        return sorted.map((d, i) => ({
          ...d,
          yoy: i === 0 ? null : (d.retail_sales / sorted[i-1].retail_sales - 1) * 100
        }))
      })
    }
  }

  const metric = showYoy ? "yoy" : "retail_sales"
  const yFmt   = showYoy ? d => `${(+d).toFixed(1)}%` : d => `$${(+d / 1e6).toFixed(0)}M`
  const yLabel  = showYoy ? "YoY % change" : "Retail Sales ($M)"

  // State YoY reference line — only meaningful in YoY mode (same scale as geos)
  let stateRef = []
  if (showYoy) {
    const stateRows = transpose(state_data)
    if (isAnnual) {
      const byYear = d3.rollup(stateRows, v => d3.sum(v, d => d.retail_sales ?? 0), d => d.year)
      const years  = [...byYear.keys()].sort()
      stateRef = years.slice(1).map(yr => ({
        year: yr,
        yoy: (byYear.get(yr) / byYear.get(yr - 1) - 1) * 100,
        label: "State"
      }))
    } else {
      stateRef = stateRows.filter(d => d.yoy != null && !isNaN(d.yoy)).map(d => ({...d, label: "State"}))
    }
  }

  return Plot.plot({
    width: 760,
    height: 360,
    color: { legend: true },
    x: { label: null, tickFormat: isAnnual ? d => String(d) : undefined },
    y: { label: yLabel, tickFormat: yFmt },
    marks: [
      Plot.lineY(data.filter(d => metric === "retail_sales" || (d[metric] != null && !isNaN(d[metric]))), {
        x: isAnnual ? "year" : (d => new Date(d.date)),
        y: metric,
        stroke: geoCol,
        tip: true,
        strokeWidth: 2
      }),
      // State YoY reference (only in YoY mode — same scale)
      ...(!showYoy ? [] : [
        Plot.lineY(stateRef, {
          x: isAnnual ? "year" : (d => new Date(d.date)),
          y: "yoy",
          stroke: "#888",
          strokeDasharray: "5,3",
          strokeWidth: 1.5
        }),
        Plot.text(stateRef.slice(-1), {
          x: isAnnual ? "year" : (d => new Date(d.date)),
          y: "yoy",
          text: "State avg",
          dy: -8,
          fontSize: 11,
          fill: "#888"
        })
      ])
    ]
  })
}
```

```{ojs}
//| title: "Latest Month: Retail Sales Ranking"

{
  const raw = geo_level === "County"
    ? transpose(county_data)
    : transpose(city_data)
  const geoCol = geo_level === "County" ? "county" : "city"

  // Most recent date
  const maxTs  = d3.max(raw, d => +new Date(d.date))
  const latest = raw.filter(d => +new Date(d.date) === maxTs && d.retail_sales != null)

  const stateTotal = transpose(state_data).find(d => d.date === maxDate)?.retail_sales ?? 1

  const sorted = latest
    .sort((a, b) => b.retail_sales - a.retail_sales)
    .slice(0, 20)
    .map(d => ({ ...d, share: d.retail_sales / stateTotal * 100 }))

  const selected = new Set(selected_geos)

  return Plot.plot({
    width: 760,
    height: 480,
    marginLeft: 130,
    x: { label: "Retail Sales", tickFormat: d => `$${(d/1e6).toFixed(0)}M` },
    y: { label: null },
    color: { scheme: "tableau10" },
    marks: [
      Plot.barX(sorted, {
        x: "retail_sales",
        y: geoCol,
        sort: { y: "-x" },
        fill: d => selected.has(d[geoCol]) ? "#2E6DA4" : "#ccc",
        tip: true
      }),
      Plot.ruleX([0])
    ]
  })
}
```

# Industry

## Row {.sidebar}

```{ojs}
viewof ind_geo_level = Inputs.radio(["State", "County", "City"], { value: "State", label: "Geography:" })

viewof ind_selected_geo = {
  // Always use Inputs.select to avoid OJS reactivity issues on input-type switch
  const opts = ind_geo_level === "State"
    ? ["(Statewide)"]
    : ind_geo_level === "County" ? counties_list : cities_list
  return Inputs.select(opts, {
    value: opts[0],
    label: "Area:",
    disabled: ind_geo_level === "State"
  })
}

viewof ind_time = Inputs.radio(["Monthly", "Annual"], { value: "Annual", label: "View:" })

viewof selected_industries = Inputs.select(
  industries_list,
  {
    value: industries_list.slice(0, 8),
    multiple: true,
    label: "Industries:",
    size: 12
  }
)
```

## Row

```{ojs}
//| title: "Retail Sales by Industry"

{
  let raw, geoCol

  if (ind_geo_level === "State") {
    raw = transpose(state_ind_data)
    geoCol = null
  } else if (ind_geo_level === "County") {
    raw = transpose(county_ind_data)
    geoCol = "county"
  } else {
    raw = transpose(city_ind_data)
    geoCol = "city"
  }

  let data = raw.filter(d =>
    (geoCol == null || d[geoCol] === ind_selected_geo) &&
    selected_industries.includes(d.industry) &&
    d.retail_sales != null
  )

  if (ind_time === "Annual") {
    const grouped = d3.rollup(data, v => ({
      year: v[0].year,
      industry: v[0].industry,
      retail_sales: d3.sum(v, d => d.retail_sales ?? 0)
    }), d => `${d.industry}_${d.year}`)
    data = [...grouped.values()]
  }

  return Plot.plot({
    width: 760,
    height: 380,
    color: { legend: true },
    x: { label: null, tickFormat: ind_time === "Annual" ? d => String(d) : undefined },
    y: { label: "Retail Sales", tickFormat: d => `$${(d/1e6).toFixed(0)}M` },
    marks: [
      Plot.lineY(data, {
        x: ind_time === "Annual" ? "year" : (d => new Date(d.date)),
        y: "retail_sales",
        stroke: "industry",
        tip: true,
        strokeWidth: 1.8
      })
    ]
  })
}
```

```{ojs}
//| title: "Industry Mix — Latest Month"

{
  let raw, geoCol

  if (ind_geo_level === "State") {
    raw = transpose(state_ind_data)
    geoCol = null
  } else if (ind_geo_level === "County") {
    raw = transpose(county_ind_data)
    geoCol = "county"
  } else {
    raw = transpose(city_ind_data)
    geoCol = "city"
  }

  const maxTs  = d3.max(raw, d => +new Date(d.date))
  const latest = raw.filter(d =>
    +new Date(d.date) === maxTs &&
    (geoCol == null || d[geoCol] === ind_selected_geo) &&
    d.industry !== "Total" &&
    d.retail_sales != null
  )

  const sorted = [...latest].sort((a, b) => b.retail_sales - a.retail_sales)

  return Plot.plot({
    width: 760,
    height: 460,
    marginLeft: 210,
    x: { label: "Retail Sales", tickFormat: d => `$${(d/1e6).toFixed(0)}M` },
    y: { label: null },
    marks: [
      Plot.barX(sorted, {
        x: "retail_sales",
        y: "industry",
        sort: { y: "-x" },
        fill: "#2E6DA4",
        fillOpacity: 0.8,
        tip: true
      }),
      Plot.ruleX([0])
    ]
  })
}
```

# Outperformance

## Row

::: {.callout-note}
**Outperformance** = a county or city whose year-over-year retail sales growth exceeds the statewide rate in the same month. Higher positive values indicate stronger-than-average growth.
:::

```{ojs}
//| title: "Controls"

viewof out_geo_level = Inputs.radio(["County", "City"], { value: "County", label: "Geography:" })
viewof out_year = Inputs.range([2017, 2026], { value: 2025, step: 1, label: "Year:" })
viewof out_industry = Inputs.select(
  ["All Industries", ...industries_list],
  { value: "All Industries", label: "Industry:" }
)
```

## Row

```{ojs}
//| title: "Excess Growth vs. State Average"

{
  // Annual YoY by geography
  const raw = out_geo_level === "County"
    ? transpose(county_data)
    : transpose(city_data)
  const geoCol = out_geo_level === "County" ? "county" : "city"

  // State annual totals for reference year and prior
  const stateByYear = d3.rollup(
    transpose(state_data),
    v => d3.sum(v, d => d.retail_sales ?? 0),
    d => d.year
  )
  const stateGrowth = (stateByYear.get(out_year) / stateByYear.get(out_year - 1) - 1) * 100

  // Geo annual totals
  const geoByYearGeo = d3.rollup(
    raw.filter(d => d.retail_sales != null),
    v => d3.sum(v, d => d.retail_sales ?? 0),
    d => `${d[geoCol]}_${d.year}`
  )

  const geos = [...new Set(raw.map(d => d[geoCol]))]

  const results = geos.flatMap(geo => {
    const cur  = geoByYearGeo.get(`${geo}_${out_year}`)     ?? null
    const prev = geoByYearGeo.get(`${geo}_${out_year - 1}`) ?? null
    if (!cur || !prev || prev === 0) return []
    const geoGrowth = (cur / prev - 1) * 100
    return [{ geo, growth: geoGrowth, excess: geoGrowth - stateGrowth, sales: cur }]
  })

  const sorted = results.sort((a, b) => b.excess - a.excess).slice(0, 25)

  return Plot.plot({
    width: 780,
    height: 540,
    marginLeft: 140,
    x: {
      label: "Excess Growth vs State (%)",
      tickFormat: d => `${d > 0 ? "+" : ""}${d.toFixed(1)}%`
    },
    y: { label: null },
    color: {
      domain: [true, false],
      range: ["#2ca02c", "#d62728"],
      legend: true,
      tickFormat: d => d ? "Outperforming" : "Underperforming"
    },
    marks: [
      Plot.barX(sorted, {
        x: "excess",
        y: "geo",
        sort: { y: "-x" },
        fill: d => d.excess >= 0,
        tip: {
          format: {
            x: d => `${d > 0 ? "+" : ""}${d.toFixed(1)}% vs state`,
            fill: false
          }
        }
      }),
      Plot.ruleX([0], { stroke: "#333", strokeWidth: 2 }),
      Plot.text(sorted.filter(d => d.excess >= 0), {
        x: "excess",
        y: "geo",
        text: d => `+${d.excess.toFixed(1)}%`,
        dx: 4,
        fontSize: 11,
        fill: "#2ca02c",
        textAnchor: "start"
      })
    ]
  })
}
```

```{ojs}
//| title: "State YoY Benchmark"

{
  const stateByYear = d3.rollup(
    transpose(state_data),
    v => d3.sum(v, d => d.retail_sales ?? 0),
    d => d.year
  )
  const stateGrowth = (stateByYear.get(out_year) / stateByYear.get(out_year - 1) - 1) * 100

  return html`<div style="padding: 1rem; background: #f0f4ff; border-radius: 8px; font-size: 1.1rem;">
    <strong>State YoY (${out_year}):</strong>
    <span style="color: ${stateGrowth >= 0 ? '#2ca02c' : '#d62728'}; font-weight: bold;">
      ${stateGrowth >= 0 ? "+" : ""}${stateGrowth.toFixed(1)}%
    </span>
    <br/>Geographies above this line are outperforming the state average.
  </div>`
}
```

# Snapshot

## Row

::: {.callout-note}
**Trailing 12-month window:** Mar 2025 – Feb 2026 vs. Mar 2024 – Feb 2025. County growth is compared to the statewide rate (+`{r} round(state_t12_total_yoy, 1)`%) over the same period. Industry rankings reflect statewide totals only.
:::

## Row

```{ojs}
//| title: "Counties: Strongest & Weakest vs. State"

{
  const raw = transpose(county_t12_data)

  // Separate small counties (prior T12 < $500M) for de-emphasis
  const N = 12  // top/bottom N to show

  const sorted = [...raw]
    .filter(d => d.yoy != null && !isNaN(d.yoy))
    .sort((a, b) => b.excess - a.excess)

  // Top N outperformers + bottom N underperformers
  const top    = sorted.slice(0, N)
  const bottom = sorted.slice(-N).reverse()
  const display = [
    ...top.map(d => ({ ...d, rank_group: "Outperforming" })),
    ...bottom.map(d => ({ ...d, rank_group: "Underperforming" }))
  ]

  // Sort for diverging bar: outperformers top, underperformers bottom
  display.sort((a, b) => b.excess - a.excess)

  return Plot.plot({
    width: 740,
    height: 560,
    marginLeft: 110,
    x: {
      label: "Excess growth vs. state T12 (%)",
      tickFormat: d => `${d > 0 ? "+" : ""}${d.toFixed(1)}%`
    },
    y: { label: null },
    color: {
      domain: ["Outperforming", "Underperforming"],
      range: ["#2ca02c", "#d62728"],
      legend: true
    },
    marks: [
      Plot.barX(display, {
        x: "excess",
        y: "county",
        sort: { y: "-x" },
        fill: "rank_group",
        fillOpacity: d => d.small_county ? 0.5 : 0.85,
        tip: {
          format: {
            x: false,
            fill: false,
            fillOpacity: false
          },
          title: d =>
            `${d.county}${d.small_county ? " (small county)" : ""}\n` +
            `T12 growth: ${d.yoy != null ? (d.yoy > 0 ? "+" : "") + d.yoy.toFixed(1) + "%" : "N/A"}\n` +
            `vs. state: ${d.excess != null ? (d.excess > 0 ? "+" : "") + d.excess.toFixed(1) + " ppts" : "N/A"}`
        }
      }),
      Plot.ruleX([0], { stroke: "#333", strokeWidth: 2 }),
      Plot.text(display.filter(d => d.excess >= 0), {
        x: "excess",
        y: "county",
        text: d => `+${d.excess.toFixed(1)}%`,
        dx: 4,
        fontSize: 10,
        fill: "#2ca02c",
        textAnchor: "start"
      }),
      Plot.text(display.filter(d => d.excess < 0), {
        x: "excess",
        y: "county",
        text: d => `${d.excess.toFixed(1)}%`,
        dx: -4,
        fontSize: 10,
        fill: "#d62728",
        textAnchor: "end"
      })
    ]
  })
}
```

```{ojs}
//| title: "Industries: Statewide Strength & Weakness"

{
  const raw = transpose(state_ind_t12_data)
    .filter(d => d.state_yoy != null && !isNaN(d.state_yoy))
    .sort((a, b) => b.state_yoy - a.state_yoy)

  // Short display labels — full names in tooltip
  const shortName = {
    "Accommodation": "Accommodation",
    "Building Material and Garden Equipment and Supplies Dealers": "Building Materials",
    "Clothing, Clothing Accessories, Shoe, and Jewelry Retailers": "Clothing & Jewelry",
    "Food and Beverage Retailers": "Food & Beverage Retail",
    "Food Services and Drinking Places": "Restaurants & Bars",
    "Furniture, Home Furnishings, Electronics, and Appliance Retailers": "Furniture & Electronics",
    "Gasoline Stations and Fuel Dealers": "Gas & Fuel",
    "General Merchandise Retailers": "General Merchandise",
    "Health and Personal Care Retailers": "Health & Personal Care",
    "Motor Vehicle and Parts Dealers": "Motor Vehicles",
    "Sporting Goods, Hobby, Musical Instrument, Book, and Miscellaneous Retailers": "Sporting Goods & Hobbies"
  }

  const display = raw.map(d => ({ ...d, label: shortName[d.industry] ?? d.industry }))

  return Plot.plot({
    width: 740,
    height: 420,
    marginLeft: 160,
    x: {
      label: "Trailing 12-month YoY growth (%)",
      tickFormat: d => `${d > 0 ? "+" : ""}${d.toFixed(1)}%`
    },
    y: { label: null },
    color: {
      domain: [true, false],
      range: ["#2ca02c", "#d62728"],
      legend: true,
      tickFormat: d => d ? "Growing" : "Declining"
    },
    marks: [
      Plot.barX(display, {
        x: "state_yoy",
        y: "label",
        sort: { y: "-x" },
        fill: d => d.state_yoy >= 0,
        fillOpacity: 0.85,
        tip: {
          format: { x: false, fill: false, fillOpacity: false },
          title: d =>
            `${d.industry}\n` +
            `T12 growth: ${(d.state_yoy > 0 ? "+" : "") + d.state_yoy.toFixed(1)}%\n` +
            `T12 sales: $${(d.sales_t12 / 1e9).toFixed(2)}B`
        }
      }),
      Plot.ruleX([0], { stroke: "#333", strokeWidth: 2 }),
      Plot.text(display.filter(d => d.state_yoy >= 0), {
        x: "state_yoy",
        y: "label",
        text: d => `+${d.state_yoy.toFixed(1)}%`,
        dx: 4,
        fontSize: 10,
        fill: "#2ca02c",
        textAnchor: "start"
      }),
      Plot.text(display.filter(d => d.state_yoy < 0), {
        x: "state_yoy",
        y: "label",
        text: d => `${d.state_yoy.toFixed(1)}%`,
        dx: -4,
        fontSize: 10,
        fill: "#d62728",
        textAnchor: "end"
      })
    ]
  })
}
```

# Area × Industry

## Row {.sidebar}

```{ojs}
viewof hm_metric = Inputs.radio(
  ["Excess vs. state industry", "Raw YoY growth"],
  { value: "Excess vs. state industry", label: "Color metric:" }
)

viewof hm_clamp = Inputs.range(
  [5, 40],
  { value: 20, step: 5, label: "Color scale cap (±%):" }
)
```

::: {.callout-note style="font-size: 0.85rem; margin-top: 1rem;"}
**Coverage:** Industry detail is available for the **11 major counties** (Adams, Arapahoe, Boulder, Denver, Douglas, El Paso, Jefferson, Larimer, Mesa, Pueblo, Weld) for 2022–present. Color scale is capped at the selected cap value; actual values appear on hover. Weld County / Sporting Goods contains an anomalous Dec 2025 spike — interpret with caution.
:::

## Row

```{ojs}
//| title: "County × Industry Growth Heatmap (Trailing 12 Months)"

{
  const raw = transpose(heatmap_data)

  // Abbreviate industry names for y-axis
  const shortInd = {
    "Accommodation": "Accommodation",
    "Building Material and Garden Equipment and Supplies Dealers": "Building Materials",
    "Clothing, Clothing Accessories, Shoe, and Jewelry Retailers": "Clothing & Jewelry",
    "Food and Beverage Retailers": "Food & Bev Retail",
    "Food Services and Drinking Places": "Restaurants & Bars",
    "Furniture, Home Furnishings, Electronics, and Appliance Retailers": "Furniture & Electronics",
    "Gasoline Stations and Fuel Dealers": "Gas & Fuel",
    "General Merchandise Retailers": "General Merchandise",
    "Health and Personal Care Retailers": "Health & Personal Care",
    "Motor Vehicle and Parts Dealers": "Motor Vehicles",
    "Sporting Goods, Hobby, Musical Instrument, Book, and Miscellaneous Retailers": "Sporting Goods & Hobbies"
  }

  const metricCol = hm_metric === "Excess vs. state industry" ? "excess" : "yoy"
  const cap       = hm_clamp  // user-controlled color scale cap

  const data = raw
    .filter(d => d[metricCol] != null && !isNaN(d[metricCol]))
    .map(d => ({
      ...d,
      ind_label:     shortInd[d.industry] ?? d.industry,
      value:         d[metricCol],
      value_clamped: Math.max(-cap, Math.min(cap, d[metricCol]))
    }))

  // Sort industry y-axis by statewide T12 YoY (strongest at top)
  const indOrder = [...d3.rollup(data, v => v[0].state_yoy, d => d.ind_label)]
    .sort((a, b) => b[1] - a[1])
    .map(d => d[0])

  const colorLabel = hm_metric === "Excess vs. state industry"
    ? `Excess growth vs. state industry (ppts, capped ±${cap}%)`
    : `YoY growth (%, capped ±${cap}%)`

  return Plot.plot({
    width:      820,
    height:     420,
    marginLeft: 165,
    marginBottom: 80,
    color: {
      type:   "diverging",
      scheme: "RdYlGn",
      domain: [-cap, cap],
      label:  colorLabel,
      legend: true
    },
    x: {
      label:      null,
      tickRotate: -35,
      domain:     [...new Set(data.map(d => d.county))].sort()
    },
    y: {
      label:  null,
      domain: indOrder
    },
    marks: [
      Plot.cell(data, {
        x:    "county",
        y:    "ind_label",
        fill: "value_clamped",
        tip: {
          format: { fill: false },
          title: d => {
            const metricLabel = hm_metric === "Excess vs. state industry"
              ? `Excess vs. state: ${(d.excess > 0 ? "+" : "") + d.excess.toFixed(1)} ppts`
              : `YoY growth: ${(d.yoy > 0 ? "+" : "") + d.yoy.toFixed(1)}%`
            const stateRef = `State industry T12: ${(d.state_yoy > 0 ? "+" : "") + d.state_yoy.toFixed(1)}%`
            return `${d.county}\n${d.industry}\n${metricLabel}\n${stateRef}`
          }
        }
      }),
      // Outline cells that are beyond the color cap (extreme values)
      Plot.cell(data.filter(d => Math.abs(d.value) > cap), {
        x:             "county",
        y:             "ind_label",
        stroke:        "#333",
        strokeWidth:   1.5,
        fill:          "none"
      })
    ]
  })
}
```

```{ojs}
//| title: "Notable Area × Industry Combinations"

{
  const raw = transpose(heatmap_data)
    .filter(d => d.excess != null && !isNaN(d.excess) && d.yoy != null && !isNaN(d.yoy))

  const shortInd = {
    "Accommodation": "Accommodation",
    "Building Material and Garden Equipment and Supplies Dealers": "Building Materials",
    "Clothing, Clothing Accessories, Shoe, and Jewelry Retailers": "Clothing & Jewelry",
    "Food and Beverage Retailers": "Food & Bev Retail",
    "Food Services and Drinking Places": "Restaurants & Bars",
    "Furniture, Home Furnishings, Electronics, and Appliance Retailers": "Furniture & Electronics",
    "Gasoline Stations and Fuel Dealers": "Gas & Fuel",
    "General Merchandise Retailers": "General Merchandise",
    "Health and Personal Care Retailers": "Health & Personal Care",
    "Motor Vehicle and Parts Dealers": "Motor Vehicles",
    "Sporting Goods, Hobby, Musical Instrument, Book, and Miscellaneous Retailers": "Sporting Goods & Hobbies"
  }

  const metricCol = hm_metric === "Excess vs. state industry" ? "excess" : "yoy"

  // Top 8 + bottom 8 county×industry combos by chosen metric
  const sorted = [...raw].sort((a, b) => b[metricCol] - a[metricCol])
  const top    = sorted.slice(0, 8).map(d => ({ ...d, group: "Standout" }))
  const bottom = sorted.slice(-8).reverse().map(d => ({ ...d, group: "Lagging" }))
  const display = [...top, ...bottom]
    .sort((a, b) => b[metricCol] - a[metricCol])
    .map(d => ({
      ...d,
      label: `${d.county} / ${shortInd[d.industry] ?? d.industry}`
    }))

  const xLabel = hm_metric === "Excess vs. state industry"
    ? "Excess growth vs. state industry (ppts)"
    : "YoY growth (%)"

  return Plot.plot({
    width:       740,
    height:      520,
    marginLeft:  230,
    x: {
      label:      xLabel,
      tickFormat: d => `${d > 0 ? "+" : ""}${d.toFixed(1)}%`
    },
    y: { label: null },
    color: {
      domain: ["Standout", "Lagging"],
      range:  ["#2ca02c", "#d62728"],
      legend: true
    },
    marks: [
      Plot.dot(display, {
        x:    metricCol,
        y:    "label",
        sort: { y: "-x" },
        r:    6,
        fill: "group",
        tip: {
          format: { fill: false, r: false },
          title: d =>
            `${d.county}\n${d.industry}\n` +
            `T12 YoY: ${(d.yoy > 0 ? "+" : "") + d.yoy.toFixed(1)}%\n` +
            `State industry T12: ${(d.state_yoy > 0 ? "+" : "") + d.state_yoy.toFixed(1)}%\n` +
            `Excess: ${(d.excess > 0 ? "+" : "") + d.excess.toFixed(1)} ppts`
        }
      }),
      Plot.ruleX([0], { stroke: "#333", strokeWidth: 1.5 }),
      Plot.text(display.filter(d => d[metricCol] >= 0), {
        x:          metricCol,
        y:          "label",
        text:       d => `${(d[metricCol] > 0 ? "+" : "") + d[metricCol].toFixed(1)}%`,
        dx:         10,
        fontSize:   10,
        fill:       "#2ca02c",
        textAnchor: "start"
      }),
      Plot.text(display.filter(d => d[metricCol] < 0), {
        x:          metricCol,
        y:          "label",
        text:       d => `${d[metricCol].toFixed(1)}%`,
        dx:         -10,
        fontSize:   10,
        fill:       "#d62728",
        textAnchor: "end"
      })
    ]
  })
}
```
